### 2023-01-25 Strydom_et_al_2023_graph_embedding_draft.tex
Deploying to gh-pages from @ PoisotLab/ms_metaweb_perspectives@347a9937e2eb5fc9f006894aadc63420e927d746 🚀

Because the metaweb represents the joint effect of functional,
phylogenetic, and macroecological processes (Morales-Castilla et al.,
2015), it holds valuable ecological information. Specifically, it
represents the ``upper bounds'' on what the composition of the local
networks, given a local species pool, can be (see \emph{e.g.} McLeod et
al., 2021); this information can help evaluate the ability of ecological
assemblages to withstand the effects of, for example, climate change
(Fricke et al., 2022). These local networks may be reconstructed given
an appropriate knowledge of local species composition and provide
information on the structure of food webs at finer spatial scales. This
has been done for example for tree-galler-parasitoid systems (Gravel et
al., 2018), fish trophic interactions (Albouy et al., 2019), tetrapod
trophic interactions (Braga et al., 2019; O'Connor et al., 2020), and
crop-pest networks (Grünig et al., 2020). In this contribution, we
highlight the power in viewing (and constructing) metawebs as
\emph{probabilistic} objects in the context of rare interactions,
discuss how a family of machine learning tools (graph embeddings and
transfer learning) can be used to overcome data limitations to metaweb
inference, and highlight how the use of metawebs introduces important
questions for the field of network ecology.

@@

It is important to recall that a metaweb is intended as a catalogue of
all potential (feasible) interactions, which is then filtered for a
given application (Morales-Castilla et al., 2015). It is therefore
important to separate the interactions that happen ``almost surely''
(repeated observational data), ``almost never'' (repeated lack of
evidence \emph{or} evidence that the link is forbidden through
\emph{e.g.} trait mis-match), and interactions with a probability that
lays somewhere in between. In a sense, that most ecological interactions
are elusive can call for a slightly different approach to sampling: once
the common interactions are documented, the effort required in
documenting each rare interaction will increase exponentially (Jordano,
2016). Recent proposals in other fields emphasize the idea that machine
learning algorithms can also act as data generators (Hoffmann et al.,
2019): high quality observational data can be used to infer core rules
underpinning network structure, and be supplemented with synthetic data
coming from predictive models trained on them, thereby increasing the
volume of information available for analysis. Indeed, Strydom, Catchen,
et al. (2021) suggested that knowing the metaweb may render the
prediction of local networks easier, because it fixes an ``upper bound''
on which interactions can exist. In this context, a probabilistic
metaweb represents an aggregation of informative priors on the
biological feasibility of interactions, which is usually hard to obtain
yet has possibly the most potential to boost our predictive ability
(Bartomeus et al., 2016).

@@

Graph (or Network) embedding (fig.~\ref{fig:embedding}) is a family of
machine learning techniques, whose main task is to learn a mapping
function from a discrete graph to a continuous domain (Arsov \& Mirceva,
2019; Chami et al., 2022). Their main goal is to learn a low dimensional
vector representations for the nodes of the graph (embeddings), such
that key properties of the graph (e.g.~local or global structures) are
retained in the embedding space (Yan et al., 2005). Ecological networks
are an interesting candidate for the widespread application of
embeddings, as they tend to possess a shared structural backbone (see
\emph{e.g.} Bramon Mora et al., 2018 for food webs), which hints at
structural invariants in empirical data; assuming that these structural
invariants are indeed widespread, they would dominate the structure of
networks, and therefore be adequately captured by the first (lower)
dimensions of an embedding, without the need to measure derived aspects
of their structure (\emph{e.g.} motifs, paths, modularity, \ldots).

Indeed, food webs are inherently low-dimensional objects, and can be
adequately represented with less than ten dimensions (Braga et al.,
2019; Eklöf et al., 2013). Simulation results by Botella et al. (2022)
suggest that there is no best method to identify architectural
similarities between networks, and that multiple approaches need to be
tested and compared to the network descriptor of interest. This matches
previous, more general results on graph embedding, which suggest that
different embedding algorithms yield different network embeddings (Goyal
\& Ferrara, 2018), calling for a careful selection of the
problem-specific approach to use. In tbl.~\ref{tbl:methods}, we present
a selection of common graph and node embedding methods, alongside
examples of their use to predict species interactions; most of these
methods rely either on linear algebra, or on pseudo-random walks on
graphs. All forms of embeddings presented in the table share the common
property of summarizing their objects into (sets of) dense feature
vectors, that capture the overall network structure, pairwise
information on nodes, and emergent aspects of the network, in a
compressed way (\emph{i.e.} with some information loss, as we later
discuss in the illustration). Node embeddings tend to focus on
maintaining pairwise relationships, while graph embeddings focus on
maintaining the network structure. Nevertheless, some graph embedding
techniques (like RDPG, see \emph{e.g.} Wu et al., 2021) will provide
usable node-level embeddings.

@@

\hypertarget{tbl:methods}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1083}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1250}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2667}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1667}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3333}}@{}}
\caption{\label{tbl:methods}Overview of some common graph embedding
approaches, by type of embedded objects, alongside examples of their use
in the prediction of species interactions. These methods have not yet
been routinely used to predict species interactions; most examples that
we identified were either statistical associations, or analogues to
joint species distribution models. \(^a\): application is concerned with
\emph{statistical} interactions, which are not necessarilly direct
biotic interactions; \(^b\):application is concerned with joint-SDM-like
approach, which is also very close to statistical associations as
opposed to direct biotic interactions. Given the need to evaluate
different methods on a problem-specific basis, the fact that a lot of
methods have not been used on network problems is an opportunity for
benchmarking and method development. Note that the row for PCA also
applies to kernel/probabilistic PCA, which are variations on the more
general method of SVD. Note further that tSNE has been included because
it is frequently used to embed graphs, including of species
associations/interactions, despite not being strictly speaking, a graph
embedding technique (see \emph{e.g.} Chami et al., 2022)}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Method
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Object
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Technique
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Reference
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Application
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Method
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Object
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Technique
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Reference
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Application
\end{minipage} \\
\midrule()
\endhead
tSNE & nodes & statistical divergence & Hinton \& Roweis (2002) & Gibb
et al. (2021); Cieslak et al. (2020) \(^a\) \\
LINE & nodes & stochastic gradient descent & Tang et al. (2015) & \\
SDNE & nodes & gradient descent & D. Wang et al. (2016) & \\
node2vec & nodes & stochastic gradient descent & Grover \& Leskovec
(2016) & \\
HARP & nodes & meta-strategy & H. Chen et al. (2017) & \\
DMSE & joint nodes & deep neural network & D. Chen et al. (2017) & D.
Chen et al. (2017) \(^b\) \\
graph2vec & sub-graph & skipgram network & Narayanan et al. (2017) & \\
RDPG & graph & SVD & Young \& Scheinerman (2007) & Poisot et al. (2021);
Dalla Riva \& Stouffer (2016) \\
GLEE & graph & Laplacian eigenmap & Torres et al. (2020) & \\
DeepWalk & graph & stochastic gradient descent & Perozzi et al. (2014) &
Wardeh et al. (2021) \\
GraphKKE & graph & stochastic differential equation & Melnyk et al.
(2020) & Melnyk et al. (2020) \(^a\) \\
FastEmbed & graph & eigen decomposition & Ramasamy \& Madhow (2015) & \\
PCA & graph & eigen decomposition & S \& Surendran (2013) & Strydom,
Catchen, et al. (2021) \\
Joint methods & multiple graphs & multiple strategies & S. Wang et al.
(2021) & \\
\bottomrule()
\end{longtable}

The popularity of graph embedding techniques in machine learning is more
than the search for structural invariants: graphs are discrete objects,
and machine learning techniques tend to handle continuous data better.
Bringing a sparse graph into a continuous, dense vector space (Xu, 2021)
opens up a broader variety of predictive algorithms, notably of the sort
that are able to predict events as probabilities (Murphy, 2022).
Furthermore, the projection of the graph itself is a representation that
can be learned; Runghen et al. (2021), for example, used a neural
network to learn the embedding of a network in which not all
interactions were known, based on the nodes' metadata. This example has
many parallels in ecology (see fig.~\ref{fig:embedding} \textbf{C}), in
which node metadata can be represented by phylogeny, abundance, or
functional traits. Using phylogeny as a source of information assumes
(or strives to capture) the action of evolutionary processes on network
structure, which at least for food webs have been well documented (Dalla
Riva \& Stouffer, 2016; Stouffer et al., 2012;
\textbf{Stouffer2007EviExi?}); similarly, the use of functional traits
assumes that interactions can be inferred from the knowledge of
trait-matching rules, which is similarly well supported in the empirical
literature (Bartomeus, 2013; Bartomeus et al., 2016;
\textbf{Gravel2013InfFoo?}). Relating this information to an embedding
rather than a list of networks measures would allow to capture their
effect on the more fundamental aspects of network structure; conversely,
the absence of a phylogenetic or functional signal may suggest that
evolutionary/trait processes are not strong drivers of network
structure, therefore opening a new way to perform hypothesis testing.

Before moving further, it is important to clarify the epistemic status
of node values derived from embeddings: specifically, they are
\emph{not} functional traits, and therefore should not be discussed in
terms of effects or responses. As per the framework of Malaterre et al.
(2019), these values neither derive from, nor result in, changes in
organismal fitness, and should therefore not be used to quantify
\emph{e.g.} functional diversity. This holds true even when there are
correlations between latent values and functional traits: although these
enable an ecological discussion of how traits condition the structure of
the network, the existence of a statistical relationship does not
elevate the latent values to the status of functional traits.

@@

In this section, we illustrate the embedding of a collection of
bipartite networks collected by Hadfield et al. (2014), using t-SVD and
RDPG (see Strydom et al., 2022 for the full details). Briefly, an RDPG
decomposes a network into two subspaces (left and right), which are
matrices that when multiplied give an approximation of the original
network. RDPG has the particularly desirable properties of being a graph
embedding technique that produces relevant node-level feature vectors,
and provides good approximations of graphs with varied structures
(Athreya et al., 2017). The code to reproduce this example is available
as supplementary material (note, for the sake of comparison, that
Strydom, Catchen, et al., 2021 have an example using embedding through
PCA followed by prediction using a deep neural network on the same
dataset). The resulting (binary) metaweb \(\mathcal{M}\) has 2131
interactions between 206 parasites and 121 hosts, and its adjacency
matrix has full rank (\emph{i.e.} it represents a space with 121
dimensions). All analyses were done using Julia (Bezanson et al., 2017)
version 1.7.2, \emph{Makie.jl} (Danisch \& Krumbiegel, 2021), and
\emph{EcologicalNetworks.jl} (Poisot et al., 2019).

@@

In fig.~\ref{fig:illustration1}, we focus on some statistical checks of
the embedding. In panel \textbf{A}, we show that the averaged \(L_2\)
loss (\emph{i.e.} the sum of squared errors) between the empirical and
reconstructed metaweb decreases when the number of dimensions (rank) of
the subspace increases, with an inflection at 39 dimensions (out of 120
initially) according to the finite differences method. As discussed by
Runghen et al. (2021), there is often a trade-off between the number of
dimensions to use (more dimensions are more computationally demanding)
and the quality of the representation. In panel \textbf{B}, we show the
increase in cumulative variance explained at each rank, and visualize
that using 39 ranks explains about 70\% of the variance in the empirical
metaweb. This is a different information from the \(L_2\) loss (which is
averaged across interactions), as it works on the eigenvalues of the
embedding, and therefore captures higher-level features of the network.
In panel \textbf{C}, we show positions of hosts and parasites on the
first two dimensions of the left and right subspaces. Note that these
values largely skew negative, because the first dimensions capture the
coarse structure of the network: most pairs of species do not interact,
and therefore have negative values. Finally in panel \textbf{D}, we show
the predicted weight (\emph{i.e.} the result of the multiplication of
the RDGP subspaces at a rank of 25) as a function of whether the
interactions are observed, not-observed, or unknown due to lack of
co-occurrence. This reveals that the observed interactions have higher
predicted weights, although there is some overlap; the usual approach to
identify potential interactions based on this information would be a
thresholding analysis, which is outside the scope of this manuscript
(and is done in the papers cited in this illustration). Because the
values returned from RDPG are not bound to the unit interval, we
performed a clamping of the weights to the unit space, showing a
one-inflation in documented interaction, and a zero-inflation in other
species pairs. This last figure crosses from the statistical into the
ecological, by showing that species pairs with no documented
co-occurrence have weights that are not distinguishable from species
pairs with no documented interactions, suggesting that (as befits a
host-parasite model) the ability to interact is a strong predictor of
co-occurrence.

@@

The first open research problem is the taxonomic and spatial limit of
the metaweb to embed and transfer. If the initial metaweb is too narrow
in scope, notably from a taxonomic point of view, the chances of finding
another area with enough related species (through phylogenetic
relatedness or similarity of functional traits) to make a reliable
inference decreases; this would likely be indicated by large confidence
intervals during estimation of the values in the low-rank space, meaning
that the representation of the original graph is difficult to transfer
to the new problem. Alternatively, if the initial metaweb is too large
(taxonomically), then the resulting embeddings would need to represent
interactions between taxonomic groups that are not present in the new
location. This would lead to a much higher variance in the starting
dataset, and to under-dispersion in the target dataset, resulting in the
potential under or over estimation of the strength of new predicted
interactions. The lack of well documented metawebs is currently
preventing the development of more concrete guidelines. The question of
phylogenetic relatedness and distribution is notably relevant if the
metaweb is assembled in an area with mostly endemic species (\emph{e.g.}
a system that has undergone recent radiation or that has remained in
isolation for a long period of time might not have an analogous system
with which to draw knowledge from), and as with every predictive
algorithm, there is room for the application of our best ecological
judgement. Because this problem relates to distribution of species in
the geographic or phylogenetic space, it can certainly be approached
through assessing the performance of embedding transfer in simulated
starting/target species pools.

The second series of problems relate to determining which area should be
used to infer the new metaweb in, as this determines the species pool
that must be used. Metawebs can be constructed by assigning interactions
in a list of species within geographic boundaries. The upside of this
approach is that information at the country level is likely to be
required for biodiversity assessments, as countries set conservation
goals at the national level (Buxton et al., 2021), and as quantitative
instruments are designed to work at these scales (Turak et al., 2017);
specific strategies are often enacted at smaller scales, nested within a
specific country (Ray et al., 2021). But there is no guarantee that
these boundaries are meaningful. In fact, we do not have a satisfying
answer to the question of ``where does a food web stop?''; the most
promising solutions involve examining the spatial consistency of network
area relationships (Fortin et al., 2021; see \emph{e.g.} Galiana et al.,
2018, 2019, 2021), which is impossible in the absence of enough
information about the network itself. This suggests that inferred
metawebs should be further downscaled to allow for \emph{a posteriori}
analyses. The methodology for metaweb downscaling is currently limited,
and it is likely that the sustained effort to characterize the spatial
dependency of food web structure will lead to more prescriptive
guidelines about the need for prediction downscaling. Recent results by
Martins et al. (2022) suggest that networks are shaped within
eco-regions, with abrupt structural transitions from an eco-region to
the next. Should this trend hold generally, this would provide an
ecologically-relevant scale at which metawebs can be downscaled and
predicted.

The final family of problems relates less to ecological methods than to
the praxis of ecological research. Operating under the context of
national divisions, in large parts of the world, reflects nothing more
than the legacy of settler colonialism, which drives a disparity in
available ecological data. Applying any embedding to biased data does
not debias them, but instead embeds these very same biases, propagating
them to the machine learning models using embeddings to make
predictions. Indeed, the use of ecological data is not an apolitical act
(Nost \& Goldstein, 2021), as data infrastructures tend to be designed
to answer questions within national boundaries (therefore placing
contingencies on what is available to be embedded), and their use often
draws upon and reinforces territorial statecraft. As per Machen \& Nost
(2021), this is particularly true when the output of ``algorithmic
thinking'' (\emph{e.g.} relying on machine learning to generate
knowledge) can be re-used for governance (\emph{e.g.} enacting
conservation decisions at the national scale). As information on species
interaction networks structure is increasingly leveraged as a tool to
guide conservation actions {[}see \emph{e.g.} recent discussions for
food-web based conservation; Eero et al. (2021); Naman et al. (2022);
Stier et al. (2017){]}, the need to appraise and correct biases that are
unwittingly propagated to algorithms when embedded from the original
data is paramount. These considerations are even more urgent in the
specific context of biodiversity data, as long-term colonial legacies
still shape taxonomic composition to this day (Lenzner et al., 2022;
Raja, 2022), and where much shorter-term changes in taxonomic and
genetic richness emerged as a consequence of environmental racism.
(Schmidt \& Garroway, 2022).

>>>>

Because the metaweb represents the joint effect of functional,
phylogenetic, and macroecological processes (Morales-Castilla et al.,
2015), it holds valuable ecological information. Specifically, it
represents the ``upper bounds'' on what the composition of the local
networks, given a local species pool, can be (see \emph{e.g.} McLeod et
al., 2021); this information can help evaluate the ability of ecological
assemblages to withstand the effects of, for example, climate change
(Fricke et al., 2022). These local networks may be reconstructed given
an appropriate knowledge of local species composition and provide
information on the structure of food webs at finer spatial scales. This
has been done for example for tree-galler-parasitoid systems (Gravel et
al., 2018), fish trophic interactions (Albouy et al., 2019), tetrapod
trophic interactions (J. Braga et al., 2019; O'Connor et al., 2020), and
crop-pest networks (Grünig et al., 2020). In this contribution, we
highlight the power in viewing (and constructing) metawebs as
\emph{probabilistic} objects in the context of rare interactions,
discuss how a family of machine learning tools (graph embeddings and
transfer learning) can be used to overcome data limitations to metaweb
inference, and highlight how the use of metawebs introduces important
questions for the field of network ecology.

@@

It is important to recall that a metaweb is intended as a catalogue of
all potential (feasible) interactions, which is then filtered for a
given application (Morales-Castilla et al., 2015). It is therefore
important to separate the interactions that happen ``almost surely''
(repeated observational data), ``almost never'' (repeated lack of
evidence \emph{or} evidence that the link is forbidden through
\emph{e.g.} trait mis-match), and interactions with a probability that
lays somewhere in between (Catchen et al., 2023). In a sense, that most
ecological interactions are elusive can call for a slightly different
approach to sampling: once the common interactions are documented, the
effort required in documenting each rare interaction will increase
exponentially (Jordano, 2016). Recent proposals in other fields relying
on machine learning approaches emphasize the idea that algorithms meant
to predict, through the assumption that they approximate the process
generating the data, can also act as data generators (Hoffmann et al.,
2019). High quality observational data can be used to infer core rules
underpinning network structure, and be supplemented with synthetic data
coming from predictive models trained on them, thereby increasing the
volume of information available for analysis. Indeed, Strydom, Catchen,
et al. (2021) suggested that knowing the metaweb may render the
prediction of local networks easier, because it fixes an ``upper bound''
on which interactions can exist. In this context, a probabilistic
metaweb represents an aggregation of informative priors on the
biological feasibility of interactions, which is usually hard to obtain
yet has possibly the most potential to boost our predictive ability
(Bartomeus, 2013; Bartomeus et al., 2016). This would represent a
departure from simple rules expressed at the network scale (\emph{e.g.}
Williams \& Martinez, 2000) to a view of network prediction based on
learning the rules that underpin interactions \emph{and} their
variability (Gupta et al., 2022).

@@

Graph (or Network) embedding (fig.~\ref{fig:embedding}) is a family of
machine learning techniques, whose main task is to learn a mapping
function from a discrete graph to a continuous domain (Chami et al.,
2022; \textbf{arsov\_network\_2019?}). Their main goal is to learn a low
dimensional vector representations of the graph (embeddings), such that
its key properties (\emph{e.g.} local or global structures) are retained
in the embedding space (Yan et al., 2005). The embedding space may, but
will not necessarily, have lower dimensionality than the graph.
Ecological networks are promising candidates for the routine application
of embeddings, as they tend to possess a shared structural backbone (see
\emph{e.g.} Bramon Mora et al., 2018), which hints at structural
invariants in empirical data. Assuming that these structural invariants
are common enough, they would dominate the structure of networks, and
therefore be adequately captured by the first (lower) dimensions of an
embedding, without the need to measure derived aspects of their
structure (\emph{e.g.} motifs, paths, modularity, \ldots).

Indeed, food webs are inherently low-dimensional objects, and can be
adequately represented with less than ten dimensions (J. Braga et al.,
2019; M. P. Braga et al., 2021; Eklöf et al., 2013). Simulation results
by Botella et al. (2022) suggests that there is no dominant method to
identify architectural similarities between networks: multiple
approaches need to be tested and compared to the network descriptor of
interest on a problem-specific basis. This matches previous results on
graph embedding, wherein different embedding algorithms yield different
network embeddings (Goyal \& Ferrara, 2018), calling for a careful
selection of the problem-specific approach to use. In
tbl.~\ref{tbl:methods}, we present a selection of common graph and node
embedding methods, alongside examples of their use to predict
interactions or statistical associations between species. These methods
rely largely on linear algebra or pseudo-random walks on graphs. All
forms of embeddings presented in the table share the common property of
summarizing their objects into (sets of) dense feature vectors, that
capture the overall network structure, pairwise information on nodes,
and emergent aspects of the network, in a compressed way (\emph{i.e.}
with some information loss, as we later discuss in the illustration).
Node embeddings tend to focus on maintaining pairwise relationships
(\emph{i.e.} species interactions), while graph embeddings focus on
maintaining the network structure (\emph{i.e.} emergent properties).
Nevertheless, some graph embedding techniques (like RDPG, see
\emph{e.g.} Wu et al., 2021) will provide high-quality node-level
embeddings while also preserving network structure.

@@

\hypertarget{tbl:methods}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1016}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1172}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2500}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1875}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3438}}@{}}
\caption{\label{tbl:methods}Overview of some common graph embedding
approaches, by type of embedded objects, alongside examples of their use
in the prediction of species interactions. These methods have not yet
been routinely used to predict species interactions; most examples that
we identified were either statistical associations, or analogues to
joint species distribution models. \(^a\): application is concerned with
\emph{statistical} interactions, which are not necessarilly direct
biotic interactions; \(^b\):application is concerned with joint-SDM-like
approach, which is also very close to statistical associations as
opposed to direct biotic interactions. Given the need to evaluate
different methods on a problem-specific basis, the fact that a lot of
methods have not been used on network problems is an opportunity for
benchmarking and method development. Note that the row for PCA also
applies to kernel/probabilistic PCA, which are variations on the more
general method of SVD. Note further that tSNE has been included because
it is frequently used to embed graphs, including of species
associations/interactions, despite not being strictly speaking, a graph
embedding technique (see \emph{e.g.} Chami et al., 2022)}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Method
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Object
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Technique
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Reference
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Application
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Method
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Object
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Technique
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Reference
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Application
\end{minipage} \\
\midrule()
\endhead
tSNE & nodes & statistical divergence & Hinton \& Roweis (2002) &
Cieslak et al. (2020) \(^a\) Gibb et al. (2021) \\
LINE & nodes & stochastic gradient descent & Tang et al. (2015) & \\
SDNE & nodes & gradient descent & D. Wang et al. (2016) & \\
node2vec & nodes & stochastic gradient descent & Grover \& Leskovec
(2016) & \\
HARP & nodes & meta-strategy & H. Chen et al. (2017) & \\
DMSE & joint nodes & deep neural network & D. Chen et al. (2017) & D.
Chen et al. (2017) \(^b\) \\
graph2vec & sub-graph & skipgram network & Narayanan et al. (2017) & \\
RDPG & graph & SVD & Young \& Scheinerman (2007) & Dalla Riva \&
Stouffer (2016); Poisot et al. (2021) \\
GLEE & graph & Laplacian eigenmap & Torres et al. (2020) & \\
DeepWalk & graph & stochastic gradient descent & Perozzi et al. (2014) &
Wardeh et al. (2021) \\
GraphKKE & graph & stochastic differential equation & Melnyk et al.
(2020) & Melnyk et al. (2020) \(^a\) \\
FastEmbed & graph & eigen decomposition & Ramasamy \& Madhow (2015) & \\
PCA & graph & eigen decomposition & Surendran (2013) & Strydom, Catchen,
et al. (2021) \\
Joint methods & multiple graphs & multiple strategies & S. Wang et al.
(2021) & \\
\bottomrule()
\end{longtable}

The popularity of graph embedding techniques in machine learning is more
than the search for structural invariants: graphs are discrete objects,
and machine learning techniques tend to handle continuous data better.
Bringing a sparse graph into a continuous, dense vector space (Xu, 2021)
opens up a broader variety of predictive algorithms, notably of the sort
that are able to predict events as probabilities (Murphy, 2022).
Furthermore, the projection of the graph itself is a representation that
can be learned; Runghen et al. (2021), for example, used a neural
network to learn the embedding of a network in which not all
interactions were known, based on the nodes' metadata. This example has
many parallels in ecology (see fig.~\ref{fig:embedding} \textbf{C}), in
which node metadata can be represented by phylogeny, abundance, or
functional traits. Using phylogeny as a source of information assumes
(or strives to capture) the action of evolutionary processes on network
structure, which at least for food webs have been well documented (M. P.
Braga et al., 2021; Dalla Riva \& Stouffer, 2016; Eklöf \& Stouffer,
2016; Stouffer et al., 2012; Stouffer et al., 2007); similarly, the use
of functional traits assumes that interactions can be inferred from the
knowledge of trait-matching rules, which is similarly well supported in
the empirical literature (Bartomeus, 2013; Bartomeus et al., 2016;
Goebel et al., 2023; Gravel et al., 2013). Relating this information to
an embedding rather than a list of networks measures would allow to
capture their effect on the more fundamental aspects of network
structure; conversely, the absence of a phylogenetic or functional
signal may suggest that evolutionary/trait processes are not strong
drivers of network structure, therefore opening a new way to perform
hypothesis testing.

Before moving further, it is important to clarify the epistemic status
of node values derived from embeddings: specifically, they are
\emph{not} functional traits, and therefore should not be discussed in
terms of effects or responses. As per the framework of Malaterre et al.
(2019), these values neither derive from, nor result in, changes in
organismal performance, and should therefore not be used to quantify
\emph{e.g.} functional diversity. This holds true even when there are
correlations between latent values and functional traits: although these
enable an ecological discussion of how traits condition the structure of
the network, the existence of a statistical relationship does not
elevate the latent values to the status of functional traits.

@@

In this section, we illustrate the embedding of a collection of
bipartite networks collected by Hadfield et al. (2014), using t-SVD and
RDPG. Briefly, an RDPG decomposes a network into two subspaces (left and
right), which are matrices that when multiplied give an approximation of
the original network. RDPG has the particularly desirable properties of
being a graph embedding technique that produces relevant node-level
feature vectors, and provides good approximations of graphs with varied
structures (Athreya et al., 2017). The code to reproduce this example is
available as supplementary material (note, for the sake of comparison,
that Strydom, Catchen, et al., 2021 have an example using embedding
through PCA followed by prediction using a deep neural network on the
same dataset). The resulting (binary) metaweb \(\mathcal{M}\) has 2131
interactions between 206 parasites and 121 hosts, and its adjacency
matrix has full rank (\emph{i.e.} it represents a space with 121
dimensions). All analyses were done using Julia (Bezanson et al., 2017)
version 1.7.2, \emph{Makie.jl} (Danisch \& Krumbiegel, 2021), and
\emph{EcologicalNetworks.jl} (Poisot et al., 2019).

@@

In fig.~\ref{fig:illustration1}, we focus on some statistical checks of
the embedding. In panel \textbf{A}, we show that the averaged \(L_2\)
loss (\emph{i.e.} the sum of squared errors) between the empirical and
reconstructed metaweb decreases when the number of dimensions (rank) of
the subspace increases, with an inflection at 39 dimensions (out of 120
initially) according to the finite differences method. As discussed by
Runghen et al. (2021), there is often a trade-off between the number of
dimensions to use (more dimensions are more computationally demanding)
and the quality of the representation. In panel \textbf{B}, we show the
increase in cumulative variance explained at each rank, and visualize
that using 39 ranks explains about 70\% of the variance in the empirical
metaweb. This is a different information from the \(L_2\) loss (which is
averaged across interactions), as it works on the eigenvalues of the
embedding, and therefore captures higher-level features of the network.
In panel \textbf{C}, we show positions of hosts and parasites on the
first two dimensions of the left and right subspaces. Note that these
values largely skew negative, because the first dimensions capture the
coarse structure of the network: most pairs of species do not interact,
and therefore have negative values. Finally in panel \textbf{D}, we show
the predicted weight (\emph{i.e.} the result of the multiplication of
the RDGP subspaces at a rank of 39) as a function of whether the
interactions are observed, not-observed, or unknown due to lack of
co-occurrence. This reveals that the observed interactions have higher
predicted weights, although there is some overlap; the usual approach to
identify potential interactions based on this information would be a
thresholding analysis, which is outside the scope of this manuscript
(and is done in the papers cited in this illustration). Because the
values returned from RDPG are not bound to the unit interval, we
performed a clamping of the weights to the unit space, showing a
one-inflation in documented interactions, and a zero-inflation in other
species pairs. This last figure crosses from the statistical into the
ecological, by showing that species pairs with no documented
co-occurrence have weights that are not distinguishable from species
pairs with no documented interactions, suggesting that (as befits a
host-parasite model) the ability to interact is a strong predictor of
co-occurrence.

@@

\hypertarget{identifying-the-properties-of-the-network-to-embed}{%
\subsection{Identifying the properties of the network to
embed}\label{identifying-the-properties-of-the-network-to-embed}}

If the initial metaweb is too narrow in scope, notably from a taxonomic
point of view, the chances of finding another area with enough related
species (through phylogenetic relatedness or similarity of functional
traits) to make a reliable inference decreases; this would likely be
indicated by large confidence intervals during estimation of the values
in the low-rank space, meaning that the representation of the original
graph is difficult to transfer to the new problem. Alternatively, if the
initial metaweb is too large (taxonomically), then the resulting
embeddings would need to represent interactions between taxonomic groups
that are not present in the new location. This would lead to a much
higher variance in the starting dataset, and to under-dispersion in the
target dataset, resulting in the potential under or over estimation of
the strength of new predicted interactions. The lack of well documented
metawebs is currently preventing the development of more concrete
guidelines. The question of phylogenetic relatedness and distribution is
notably relevant if the metaweb is assembled in an area with mostly
endemic species (\emph{e.g.} a system that has undergone recent
radiation or that has remained in isolation for a long period of time
might not have an analogous system with which to draw knowledge from),
and as with every predictive algorithm, there is room for the
application of our best ecological judgement. Because this problem
relates to distribution of species in the geographic or phylogenetic
space, it can certainly be approached through assessing the performance
of embedding transfer in simulated starting/target species pools.

\hypertarget{identifying-the-scope-of-the-prediction-to-perform}{%
\subsection{Identifying the scope of the prediction to
perform}\label{identifying-the-scope-of-the-prediction-to-perform}}

The area used to infer the new metaweb in determines the species pool
that must be used to perform this embedding. Metawebs can be constructed
by assigning interactions in a list of species within geographic
boundaries. The upside of this approach is that information at the
country level is likely to be required for biodiversity assessments, as
countries set conservation goals at the national level (Buxton et al.,
2021), and as quantitative instruments are designed to work at these
scales (\textbf{turak\_measuring\_2017?}); specific strategies are often
enacted at smaller scales, nested within a specific country (Ray et al.,
2021). But there is no guarantee that these boundaries are meaningful.
In fact, we do not have a satisfying answer to the question of ``where
does an ecological network stop?''. Recent results by Martins et al.
(2022) suggest that networks are shaped within eco-regions, with abrupt
structural transitions from an eco-region to the next. Should this trend
hold generally, this would provide an ecologically-relevant scale at
which metawebs can be downscaled and predicted. Other solutions leverage
network-area relationships to identify areas in which networks are
structurally similar (see \emph{e.g.} Fortin et al., 2021; Galiana et
al., 2022, 2018), which requires ample pre-existing information about
the network. This suggests that inferred metawebs should be further
downscaled to allow for \emph{a posteriori} analyses; Llewelyn et al.
(2022) provide compelling evidence for the fact that when a place is
rich in endemic species, even at smaller spatial scales, the transfer of
information about interactions becomes more challenging, re-emphasizing
the need to identify guidelines for when and where network predictions
can be mapped.

\hypertarget{minding-legacies-shaping-ecological-datasets}{%
\subsection{Minding legacies shaping ecological
datasets}\label{minding-legacies-shaping-ecological-datasets}}
